A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-08-18 tex/sec6-formalisation.tex
[ tex ] Fix a typo
[ tex ] Fix an alignment issue

As shown in \Cref{fig:formal-bibisig}, we define \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}

@@

At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}}, similarly to how we have defined simple types and raw terms.
This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{56}{@{}>{\hspre}l<{\hspost}@{}}%
\column{62}{@{}>{\hspre}l<{\hspost}@{}}%
\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
\column{85}{@{}>{\hspre}l<{\hspost}@{}}%
\column{89}{@{}>{\hspre}l<{\hspost}@{}}%
\column{108}{@{}>{\hspre}l<{\hspost}@{}}%
\column{113}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[32]%
\>[32]{}\hspace{10em}{}\<[56]%
\>[56]{}\Varid{\char95 ↑\char95 }\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇐}\;\Conid{A}\;{}\<[E]%
\\[\blanklineskip]%
\>[9]{}\Varid{\char95 ∋\char95 }\;{}\<[15]%
\>[15]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
\>[56]{}\Varid{op}\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;{}\<[85]%
\>[85]{}\Varid{rs}\;{}\<[89]%
\>[89]{}\phantom{[}{}\<[108]%
\>[108]{}\Varid{d}\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;{}\<[89]%
\>[89]{}[\mskip1.5mu \;{}\<[108]%
\>[108]{}\Varid{d}\;\mskip1.5mu]\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}

@@

Compared to an induction proof tailored for a specific system, the size of a generic induction proof remains the same, regardless of the signature.
This makes it manageable even in a proof assistant like \Agda, which lacks extensive tactics for automating theorem proving.
Hence, we achieve more by doing less.

>>>>

As shown in \Cref{fig:formal-bibisig}, we define \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{constructor}\;\Varid{ι}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{constructor}\;\Varid{sigd}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}

@@

At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}}, similarly to how we have defined simple types and raw terms.
This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
{\small\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{56}{@{}>{\hspre}l<{\hspost}@{}}%
\column{62}{@{}>{\hspre}l<{\hspost}@{}}%
\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
\column{85}{@{}>{\hspre}l<{\hspost}@{}}%
\column{89}{@{}>{\hspre}l<{\hspost}@{}}%
\column{108}{@{}>{\hspre}l<{\hspost}@{}}%
\column{113}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{var}\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[32]%
\>[32]{}\hspace{10em}{}\<[56]%
\>[56]{}\Varid{\char95 ↑\char95 }\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
\>[71]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
\>[56]{}\Varid{op}\;{}\<[62]%
\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;{}\<[85]%
\>[85]{}\Varid{rs}\;{}\<[89]%
\>[89]{}\phantom{[}{}\<[108]%
\>[108]{}\Varid{d}\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;{}\<[89]%
\>[89]{}[\mskip1.5mu \;{}\<[108]%
\>[108]{}\Varid{d}\;\mskip1.5mu]\;{}\<[113]%
\>[113]{}\Conid{A}\;{}\<[E]%
\\
\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}

@@

Each construct in an induction proof tailored for a specific system such as \PCF has to be scrutinised, so its size increases as the system in question has more constructs.
On the other hand, the size of a generic induction proof remains the same and is manageable in a proof assistant that lacks extensive tactics for automating theorem proving, like \Agda.
We have achieved more by doing no ad hoc proofs.
